Qualitative Probabilistic Programming

### 2015-06-05 paper/paper.tex
paper progress

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\usepackage{nips15submit_e,times}
\setlength{\emergencystretch}{10em}

@@

  Quipp is implemented as a library for webppl programs.  Webppl [TODO cite] is a probabilistic programming language
  implemented in Javascript. Quipp programs are written as webppl programs that have access to special functions.

  Here is an example of a Quipp program to cluster 2d points into 3 clusters:
  \begin{verbatim}
var nclusters = 3;
var dim = 2;

testParamInference(function(randFunction) {
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(Categorical(nclusters), pointType);
  return function() {
    repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      observe('point' + i, getPoint, cluster);
    });
  };
});
  \end{verbatim}

  It is written as a generative model producing the observations.  Notice that
  we declared two types (\texttt{PointType} and \texttt{ClusterType}) and one
  random function \texttt{getPoint}).  Type annotations are necessary for random
  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
  points are 2-dimensional, and the type \texttt{Categorical(3)} expresses the
  fact that there are 3 possible clusters (so a cluster may be either 0, 1, or
  2).    We assume the distribution over clusters is uniform, but we do not know
  the distribution of points in each cluster.  We will fill in the
  \texttt{getPoint} function with a learned function that will take a random
  sample from the given cluster.  The \texttt{observe} function allows us to
  observe data; here it says that the observation named \texttt{point3} is equal
  to the result of calling \texttt{getPoint(cluster)}, where \texttt{cluster} is
  cluster number 3.

@@

  Note that, because the two clusters are forced to have the same variance, they do not fit the data well,
  since the data has a different shape in each location.  To fix this problem, we can substitute the following
  model:
  \begin{verbatim}
testParamInference(function() {
  var pointType = Vector(dim, Double);
  var getPointFunctions = repeat(nclusters, function(i) {
    return randFunction(pointType);
  });
  return function() {
    return repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      return observe('point' + i, getPointFunctions[cluster]);
    });
  };
});
  \end{verbatim}
  Using this model, we get the following clusters:
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
  \end{center}
  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, or 0.1058 per point, which means it
  fits the data better.

@@

    For unknown functions, the family of random functions used is a kind
    of generalized linear model.  We assume that the distribution
    of the function's return value
    is some exponential family whose natural
    parameters are determined from the arguments:

@@

    To determine $\eta(x)$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.  Features are the same as sufficient
    statistics for the categorical distribution, but while both $X$ and $X^2$ are
    sufficient statistics for the Gaussian distribution, only $X$ is a feature.

    Let $\psi(x)$ be the features of $x$.  Then
    $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$

@@

@@

testParamInference(function(randFunction) {
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(Categorical(nclusters), pointType);
  return function() {
    return repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      return observe('point' + i, getPoint, cluster);
    });
  };
});
\end{verbatim}
}

@@

testParamInference(function(randFunction) {
  var class0Features = randFunction(featuresType);
  var class1Features = randFunction(featuresType);

  var getFeatures = randFunction(Categorical(nclusters), pointType);
  return function() {
    repeat(n, function(i) {
      var whichClass = randomInteger(2);
      if (whichClass == 0) {
        observe('features' + i, class0Features);
      } else {
        observe('features' + i, class1Features);
      }
    });
  };
});
\end{verbatim}
}

@@

testParamInference(function(randFunction) {
  var factorType = Vector(nfactors, Double);
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(factorType, pointType);
  var getFactors = function() {
    return repeat(nfactors, function() { return gaussian(0, 1); });
  };
  return function() {
    return repeat(n, function(i) {
      var factors = getFactors();
      return observe('point' + i, getPoint, factors);
    });
  };
});
\end{verbatim}
}

@@

testParamInference(function(randFunction) {
  var stateType = Categorical(nstates);
  var obsType = Categorical(nobs);
  var transFun = randFunction(stateType, stateType);
  var obsFun = randFunction(stateType, obsType);

  var observeStates = function(sampIndex, startState, i) {
    if (i == chainLength) {
      return [];
    } else {
      observe('obs' + sampIndex + '_' + i, obsFun, startState);
      return [startState].concat(observeStates(sampIndex, transFun(startState), i+1));
    }
  };
  return function() {
    return repeat(n, function(sampIndex) {
      return observeStates(sampIndex, randomInteger(nstates), 0);
    });
  };
});
\end{verbatim}
}

@@

testParamInference(function(randFunction) {
  var classToWord = randFunction(ClassType, WordType);
  return function() {
    return repeat(nDocuments, function(docIndex) {
      var whichClass = randomInteger(nClasses);
      var nWords = observe('len' + docIndex, randomInteger, maxWordsPerDocument);
      repeat(nWordsPerDocument, function(wordIndex) {
        observe('word' + doc + '_' + wordIndex, classToWord, whichClass);
      });
      return whichClass;
    });
  };
});
\end{verbatim}
}

@@

>>>>

\documentclass{article}
\usepackage{cite}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\usepackage{nips15submit_e,times}
\setlength{\emergencystretch}{10em}

@@

  Quipp is implemented as a library for webppl programs.  Webppl \cite{dippl} is a probabilistic programming language
  that is similar to Javascript but also contains features for generating random values, conditioning on values,
  and estimating expectations. Quipp programs are written as webppl programs that have access to additional special functions.

  Here is an example of a Quipp program to cluster 2d points into 2 clusters:
  \begin{verbatim}
var n = 100;

var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPoint = randFunction(Categorical(2), pointType);
return function() {
  repeat(n, function(i) {
    var cluster = randomValue(Cluster);
    observe(getPoint, cluster);
  });
};
  \end{verbatim}

  It is written as a generative model for the observations.  Notice that
  we declared two types (\texttt{Point} and \texttt{Cluster}) and one
  random function \texttt{getPoint}).  Type annotations are necessary for random
  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
  points are 2-dimensional, and the type \texttt{Categorical(2)} expresses the
  fact that there are 2 possible clusters (so a cluster may be either 0 or 1).
  We use the \texttt{randomValue} function to generate a random uniform cluster. We do not know
  the distribution of points in each cluster, represented by the fact that \texttt{getPoint}
  is an unknown random function.  We will fill in the
  \texttt{getPoint} function with a learned function that will take a random
  sample from the given cluster.  The \texttt{observe} function allows us to
  observe data; here it says that the observation named \texttt{point3} is equal
  to the result of calling \texttt{getPoint(cluster)}, where \texttt{cluster} is
  cluster number 3.

@@

  Note that, because the two clusters are forced to have the same variance, they do not fit the data well,
  since the data has a different shape in each location.  To fix this problem, we can substitute the following
  model, which uses a separate random function for each cluster:
  \begin{verbatim}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPointFunctions = [randFunction(Point), randFunction(Point)];
return repeat(n, function(i) {
  var cluster = randomValue(Cluster);
  observe('point' + i, getPointFunctions[cluster]);
});
\end{verbatim}
  Using this model, we get the following clusters:
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
  \end{center}
  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, or 0.1058 per point, so it assigns significantly higher likelihood to the data.

  Note that the clusters are still restricted to have $x$ and $y$ be independent.  To allow these coordinates
  to be correlated within a cluster, we can use the following model:

\begin{verbatim}

var Cluster = Categorical(2);
var getX = [randFunction(Double), randFunction(Double)];
var getY = [randFunction(Double, Double), randFunction(Double, Double)];

return function() {
  return repeat(n, function(i) {
    var cluster = randomInteger(nclusters);
    var x = observe(getX[cluster]);
    var y = observe(getY[cluster], x);
  });
};
\end{verbatim}

  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend on the $x$ coordinate).  This allows $x$ and $y$ to be correlated, improving fit:

  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_dep.png}
  \end{center}

  This model is estimated to assign probability $e^{-269}$ to the data, or 0.166 per point, a large improvement from the previous model.

@@

    For unknown functions, the family of random functions used is a kind
    of generalized linear model.  We assume that the distribution
    of the function's return value $y$
    is some exponential family whose natural
    parameters are determined from the arguments $x$:

@@

    To determine $\eta(x)$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
    and features are both $[X = 1], [X = 2] ..., [X=n - 1]$.
    The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.

    Let $\psi(x)$ be the features of $x$.  Then
  $$\eta(x) = \mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}$$
    $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$

@@

    \begin{tabular}{| c | c | c | c|}
      \hline
      T & $\phi_T(X)$ & $\psi_T(X)$ & Random function class \\

      \hline
    \texttt{Double} & $\begin{bmatrix} X \\ X^2 \end{bmatrix}$ & $\begin{bmatrix} X \end{bmatrix}$ & Linear regression with Gaussian noise \\

      \hline
    \texttt{Categorical(n)} & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $\begin{bmatrix} [X = 1] \\ [X = 2] \\ ...\\ [X = n-1] \end{bmatrix}$ & $n$-class logistic regression \\

      \hline
    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
      Independent prediction of each component
      \\
      \hline
    \end{tabular}

@@

var pointType = Vector(dim, Double);
var getPoint = randFunction(Categorical(nclusters), pointType);
return function() {
  return repeat(n, function(i) {
    var cluster = randomInteger(nclusters);
    return observe('point' + i, getPoint, cluster);
  });
};
\end{verbatim}
}

@@

var class0Features = randFunction(featuresType);
var class1Features = randFunction(featuresType);

var getFeatures = randFunction(Categorical(nclusters), pointType);
return function() {
  repeat(n, function(i) {
    var whichClass = randomInteger(2);
    if (whichClass == 0) {
      observe('features' + i, class0Features);
    } else {
      observe('features' + i, class1Features);
    }
  });
};
\end{verbatim}
}

@@

var factorType = Vector(nfactors, Double);
var pointType = Vector(dim, Double);
var getPoint = randFunction(factorType, pointType);
var getFactors = function() {
  return repeat(nfactors, function() { return gaussian(0, 1); });
};
return function() {
  return repeat(n, function(i) {
    var factors = getFactors();
    return observe('point' + i, getPoint, factors);
  });
};
\end{verbatim}
}

@@

var stateType = Categorical(nstates);
var obsType = Categorical(nobs);
var transFun = randFunction(stateType, stateType);
var obsFun = randFunction(stateType, obsType);

var observeStates = function(sampIndex, startState, i) {
  if (i == chainLength) {
    return [];
  } else {
    observe('obs' + sampIndex + '_' + i, obsFun, startState);
    return [startState].concat(observeStates(sampIndex, transFun(startState), i+1));
  }
};

return function() {
  return repeat(n, function(sampIndex) {
    return observeStates(sampIndex, randomInteger(nstates), 0);
  });
};
\end{verbatim}
}

@@

var classToWord = randFunction(ClassType, WordType);
return function() {
  return repeat(nDocuments, function(docIndex) {
    var whichClass = randomInteger(nClasses);
    var nWords = observe('len' + docIndex, randomInteger, maxWordsPerDocument);
    repeat(nWordsPerDocument, function(wordIndex) {
      observe('word' + doc + '_' + wordIndex, classToWord, whichClass);
    });
    return whichClass;
  });
};
\end{verbatim}
}

@@

\bibliography{bibliography}{}
\bibliographystyle{plain}
